`State and Parameter Estimation for Spatio-Temporal Neural Fields' or `A Data Driven Framework for Patient-Specific Neural Field Modelling'

### 2010-06-22 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
general paper edits, to word things more clearly

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite-dimensional state-space model. Spatial frequency analysis methods are introduced for model selection by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch Tung Striebel Smoother for state estimation and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic dynamics with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

\subsection{Derivation of Finite Dimensional State-Space Model}\section{Sect:ReducedModelDerivation}

@@

\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
% \subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT.pdf}}\\
\caption{Cross-section of connectivity kernel estimate. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the $95~\%$ confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
\end{figure}
The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over iterations) of the root mean square error (MRMSE) (over space). The RMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}, with the $95~\%$ confidence interval. At each time point the error is consistently around 0.5, which approximately $7.1$~\% of the field voltage range, which is $(-3.29,3.77)$~mV. The error in the reconstructed field is mostly due to the loss of high spatial frequency information due to the basis function decomposition as illustrated in Figure~\ref{fig:FieldEstimate}. The figure shows a snapshot at a single time instant of the true neural field, the estimated field and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Figure~\ref{fig:FieldEstimate}(b) that the estimated field contains the basic structure of the true field (Figure~\ref{fig:FieldEstimate}(a)), but is missing the higher-frequency detail.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region).} 
\label{fig:RMSE}
   \end{figure}

@@

\section{Discussion}\label{DiscussionSection}
This paper has presented a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. The results indicate good estimation performance, where the mean parameter estimates are in good agreement to the actual values. The estimates of the connectivity kernel parameters show relatively broad distributions, which is expected considering the system is driven by an unknown input, modelled as a random disturbance and the observations are corrupted by noise. To overcome the effect of noise, an evoked-response paradigm should be considered when using electrophysiological recordings. Following such a paradigm, data can be epoched into a sufficient number of realisations, where each realisation is a response to a stimuli, such that a distribution of the parameter estimates can be formed.

A significant contribution of this paper is the derivation of the basis function decomposition of the neural field model, as described in Section~\ref{Sect:ReducedModelDerivation}. Through the derivation, a finite-dimensional state-space model is formed that describes a continuous neural field. Previous work from Schiff and Sauer~\ref{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter for state-estimation of a discrete (in space and time) nonlinear spatial-temporal systems. Within their framework, each discrete spatial location forms an element of the state vector to be estimated with the UKF. Therefore, a limitation of this framework lies in an increase in dimensionality with increases in  the spatial resolution and in the physical size of the neural field. Increases in dimensionality lead to increases in the uncertainty the estimated state, and problems associated with computational complexity. This paper builds on this previous study, where the estimator is constructed to describe a continuous field. The dimensionality of the state vector is independent of the resolution of a spatial discretisation and is governed by the inherent spatial properties of the field. The frequency analysis methods provide a rigorous framework for the placement and the number of the basis functions required for the decomposition. This enables estimation of a larger field, with a lower dimensional state vector. The basis function decomposition also enabled the coupling to be described by an inner product rather than a spatial convolution. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

The frequency analysis and model selection tools presented in Section~\ref{SpectralAnalysisSection} specifies the sensor arrangement that is required for capturing the dominant cortical dynamics. Importantly, these new tools can also be used to better design electrophysiological recording techniques to avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}.

To the authors' best knowledge, this is the first paper to propose a model-based estimation framework for intracortical connectivity from electrophysiological data. The importance of the estimation framework can be illustrated by considering that the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion of water in grey matter~\cite{Assaf2008}.

@@

\section{Conclusion}
This paper presents a novel framework for state and parameter estimation of spatio-temporal neural fields to create patient specific models. A basis function decomposition reduces the infinite dimensional neural field (continuous space) to a finite dimensional state-space model. The frequency analysis methods presented in this paper provide a systematic procedure for specifying the sensor and basis function arrangement required to describe the significant neuro-dynamics by the reduced neural field model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch Tung Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. We have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic dynamics using the proposed methodology.

>>>>

\begin{abstract}
	This paper presents a framework for creating patient-specific neural field models from electrophysiological data. The Wilson and Cowen or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite-dimensional state-space model. Spatial frequency analysis methods are introduced for model selection by systematically specifying the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch-Tung-Striebel smoother for state estimation and a least squares algorithm for parameter estimation. The results show that it is possible to reconstruct the neural field and estimate intracortical connectivity and synaptic dynamics with the proposed framework. The results also illustrate the loss of high spatial frequency information as the cost incurred by the model reduction procedure. This framework provides a link between patient-specific neurophysiological data and theoretical neural fields models. This link may lead to greater understanding of cortical dynamics at the meso/macroscopic scale where diseases such as epilepsy are manifested.   
\end{abstract}

@@

\subsection{Derivation of Finite Dimensional State-Space Model}\label{Sect:ReducedModelDerivation}

@@

\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
% \subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT.pdf}}\\
\caption{Cross-section of connectivity kernel estimate. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the $95~\%$ confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
\end{figure}
The accuracy of the state estimates was evaluated by comparing the field reconstruction to the true field using the mean (over iterations) of the root mean square error (MRMSE) (over space). The MRMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}, with the $95~\%$ confidence interval. At each time point the error is consistently around 0.5~mV, which approximately $7.1$~\% of the field voltage range, which is $(-3.29,3.77)$~mV. The error in the reconstructed field is mostly due to the loss of high spatial frequency information due to the basis function decomposition as illustrated in Figure~\ref{fig:FieldEstimate}. The figure shows a snapshot at a single time instant of the true neural field, the estimated field and the corresponding error, which is the difference between the true and estimated neural fields. It can be seen in Figure~\ref{fig:FieldEstimate}(b) that the estimated field contains the basic structure of the true field (Figure~\ref{fig:FieldEstimate}(a)), but is missing the higher-frequency detail.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region).} 
\label{fig:RMSE}
   \end{figure}

@@

\section{Discussion}\label{DiscussionSection}
This paper has presented a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. The results indicate good estimation performance, where the mean parameter estimates are in good agreement with the actual values. The estimates of the connectivity kernel parameters show relatively broad distributions, which is expected considering the system is driven by an unknown input, that is modelled as a random disturbance, and the observations are corrupted by noise. To overcome the effect of noise, an evoked-response paradigm should be considered when applying the estimation algorithm to electrophysiological recordings. Following such a paradigm, data can be epoched into a sufficient number of trials (realisations), where each trial is a response to a stimulus, such that a distribution of the parameter estimates can be formed.

A significant contribution of this paper is the derivation of the basis function decomposition of the neural field model, as described in Section~\ref{Sect:ReducedModelDerivation}. Through the derivation, a finite-dimensional state-space model is formed that approximates a continuous neural field. Previous work from Schiff and Sauer~\cite{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter for state estimation for discrete (in space and time) nonlinear spatio-temporal systems. Within their framework, each discrete spatial location forms an element of the state vector to be estimated with the UKF. A limitation of this framework lies in an increase in dimensionality with increases in  the spatial resolution and in the physical size of the neural field. Increases in dimensionality lead to increases in uncertainty of the estimated state and problems associated with computational complexity. This paper builds on this previous study, where the estimator is constructed to approximate a continuous field and the dimensionality of the state vector is independent of the resolution of a spatial discretisation and is governed by the inherent spatial properties of the field. The frequency analysis methods presented in Section~\ref{SpectralAnalysisSection} provide a rigorous framework for the placement and the number of the basis functions required for the decomposition. This enables estimation of a larger field, with a lower dimensional state vector. The basis function decomposition also enabled the coupling to be described by an inner product rather than a spatial convolution. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

The frequency analysis and model selection tools presented in Section~\ref{SpectralAnalysisSection} specify the sensor arrangement that is required for capturing the dominant cortical dynamics. Importantly, these new tools can also be used to design electrophysiological recording systems to avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}.

To the authors' best knowledge, this is the first paper derive a continuum neural field model-based estimation framework for intracortical connectivity from electrophysiological data. The importance of the estimation framework can be illustrated by considering that the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion of water in grey matter~\cite{Assaf2008}.

@@

\section{Conclusion}
This paper presents a novel framework for state and parameter estimation of spatio-temporal neural fields to create patient specific models. A basis function decomposition reduces the infinite dimensional neural field (continuous space) to a finite dimensional state-space model. The frequency analysis methods presented in this paper provide a systematic procedure for specifying the sensor and basis function arrangement required to describe the significant neuro-dynamics by the reduced neural field model. The level of detail of the estimated field is related to the basis functions decomposition, providing a flexible framework where a trade-off can be made between physiological detail and computational complexity. Using the state-space model, the unscented Rauch-Tung-Striebel smoother was applied for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation. We have demonstrated that it is theoretically possible to estimate intracortical connectivity and the synaptic dynamics using the proposed methodology.
